Data Analytics Introduction Using Python Training

### 2024-11-14 MsaPythonDataAnalytics.qmd
'update3'

@@

>>>>

data.sort_values(by = 'k1') #sort

**Sample Cleaning**

# Reading data using pandas
df = pd.read_csv("https://rcs.bu.edu/examples/python/DataAnalysis/Salaries.csv")
# List first 5 records
df.head()

#Select column

df['sex']
df.sex

# #Group data using rank
df_rank = df.groupby(["rank"])

df_rank.head()

# #Calculate mean value for each numeric column per each group
df_rank.mean()

# Once groupby object is created we can calculate various statistics for each group:

#Calculate mean salary for each professor rank:
df.groupby('rank')[['salary']].mean()

# Note: If single brackets are used to specify the column (e.g. salary), then the output is Pandas Series object.
# When double brackets are used the output is a Data Frame

#Calculate mean salary for each professor rank:
df.groupby(['rank'], sort=False)[['salary']].mean()

# subset the rows in which the salary value is greater than $120K: 
df_sub = df[df['salary'] > 120000]
df_sub.head()

#Select only those rows that contain female professors:
df_f = df[df['sex'] == 'Female']

#Selecting rows
df[0:10]

#Select rows by their labels:
df_sub.loc[10:20,['rank','sex','salary']]

#Select rows by their labels:
df_sub.iloc[10:20,[0, 3, 4, 5]]

#We can sort the data using 2 or more columns:
df_sorted = df.sort_values( by =['service', 'salary'], ascending = [True, False])
df_sorted.head(10)

**Sample 2***

# Read a dataset with missing values
flights = pd.read_csv("https://rcs.bu.edu/examples/python/DataAnalysis/flights.csv")

# Select the rows that have at least one missing value
flights[flights.isnull().any(axis=1)].head()

#There are a number of methods to deal with missing values in the data frame:

#df.dropna(): drop missing observations
#df.dropna(how = "all"):  drop observtions where all cells is NA
#df.dropna(axis = 1,how = "all"): drop column if all the values are missing
#df.dropna(thresh = 5): Drop rows that contain less than 5 non-missing values
#df.fillna(0): Replace missing values with zeros
#df.isnull(): returns True if the value is missing
#df.notnull(): Returns True for non-missing values

#### Aggregation Functions in Pandas

Aggregation - computing a summary statistic about each group, i.e.
compute group sums or means
compute group sizes/counts

Common aggregation functions:

min, max
count, sum, prod
mean, median, mode, mad
std, var

flights[['dep_delay','arr_delay']].agg(['min','mean','max'])

flights.groupby('origin').count()

flights['origin'].value_counts()

@@

### Bonus material

**Collapsing data into categories**
Map categories to fewer ones: reducing categories in categorical column.

# Create mapping dictionary and replace
mapping = {'Microsoft':'ComputerOS', 'MacOS':'ComputerOS', 'Linux':'ComputerOS',
'IOS':'MobileOS', 'Android':'MobileOS'}
devices['operating_system'] = devices['operating_system'].replace(mapping)
devices['operating_system'].unique()

# Replace "+" with "00"
phones["Phone number"] = phones["Phone number"].str.replace("+", "00")
phones

# Replace "-" with nothing
phones["Phone number"] = phones["Phone number"].str.replace("-", "")
phones

# Replace phone numbers with lower than 10 digits to NaN
digits = phones['Phone number'].str.len()
phones.loc[digits < 10, "Phone number"] = np.nan
phones

# Replace letters with nothing
phones['Phone number'] = phones['Phone number'].str.replace(r'\D+', '')
phones.head()

# Treating date data
birthdays['Birthday'] = birthdays['Birthday'].dt.strftime("%d-%m-%Y")
birthdays.head()

sum_classes = flights[['economy_class', 'business_class', 'first_class']].sum(axis = 1)
passenger_equ = sum_classes == flights['total_passengers']
# Find and filter out rows with inconsistent passengers
inconsistent_pass = flights[~passenger_equ]
consistent_pass = flights[passenger_equ]

# Drop missing values
airquality_dropped = airquality.dropna(subset = ['CO2'])
airquality_dropped.head()

#Replacing with statistical measures
co2_mean = airquality['CO2'].mean()
airquality_imputed = airquality.fillna({'CO2': co2_mean})
airquality_imputed.head()

# Multiple grouped summaries
dogs.groupby("color")["weight_kg"].agg([min, max, sum])

# Grouping by multiple variables
dogs.groupby(["color", "breed"])["weight_kg"].mean()

dogs.groupby(["color", "breed"])[["weight_kg", "height_cm"]].mean()

# Group by to pivot table
dogs.pivot_table(values="weight_kg", index="color")

# Different statistics
import numpy as np
dogs.pivot_table(values="weight_kg", index="color", aggfunc=np.median)

# Multiple statistics
dogs.pivot_table(values="weight_kg", index="color", aggfunc=[np.mean, np.median])

dogs.pivot_table(values="weight_kg", index="color", columns="breed")

# Filling missing values in pivot tables
dogs.pivot_table(values="weight_kg", index="color", columns="breed", 
fill_value=0)

# Summing with pivot tables
dogs.pivot_table(values="weight_kg", index="color", columns="breed",
fill_value=0, margins=True)
